---
title: "Samplt Storage Comparison"
output: pdf_document
author: "Jacob Westaway"
date: "Last updated on `r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/Jacob/Desktop/Jacob_Uni/Data/Long_term_effects_of_probiotics')
```

# Data exploration and univariate analysis.


## About

This script explores the differences and similarities between sample storage types, and uses ordinations and EnvFit/PERMANOVA analysis (Vegan package) to determine if storage of faecal samples with the OMNIGene collection kit differed from the gold standard freezing at -80. The EnvFit function was applied to an ordination using NMDS that was based on a Bray-Curtis dissimilarity matrix calculated from data normalised with Total Sum Scaling, and the PERMANOVA to an ordination using PCoA that was also based on Bray-Curtis. The significance of both was based on 10,000 permutations and was transformed based on the Benjamini-Hochberg (BH) procedure.

## Load required packages. 

```{r, warning=F, message=F, results='hide'}
sapply(c("BiocManager", "ggplot2", "ggforce", "vegan", "knitr", "dplyr", 
         "phyloseq", "phyloseqGraphTest", "igraph", "ggnetwork", "nlme", 
         "reshape2", "tidyverse", "plyr", "DESeq2", "sjPlot", "ggpubr", 
         "gridExtra", "grid", "gtable", "lazyeval", "phyloseq", "tidyverse", 
         "knitr", "lme4", "emmeans", "MuMIn", "aods3", "sjPlot", "car"), 
       require, character.only = TRUE)
```

## Subset data to sample storage comparison

```{r, warning=F, message=F}
ps2_storage <- subset_samples(ps2, Study == "2") %>% 
  subset_samples(URN != "957652") %>% 
  subset_samples(URN != "648892")
ps3_storage <- subset_samples(ps3, Study == "2") %>% 
  subset_samples(URN != "957652") %>% 
  subset_samples(URN != "648892")
ps4_storage <- subset_samples(ps4, Study == "2") %>% 
  subset_samples(URN != "957652") %>% 
  subset_samples(URN != "648892")
```

## Beta diversity
 - Use distance and ordination methods to explore the relationship between metadata.
 - We calculate the distances using pruned, transformed (TSS) and non-agglomerated data.

```{r, warning=F, message=F, eval=F}
ps2_storage <- ps2_storage %>%
        transform_sample_counts(function(x) x / sum(x))
```
 
 - We can then create distance matrices and plots for this data subset using several methods:
 - e.g. bray-curtis or weighted unifrac distances with PCoA, NMDS, etc.
 - Define a function that ordindates the previously transformed data, extarcts the eigenvalues, and creates a dissimilarity plot.
 - Extract eigenvalues from ordination.

```{r, fig.cap="PCoA plot of Bray-Curtis distances coloured by date.", warning=F, message=F}
ordination_plots <- function(filtered_ps, variable, variable_2, vis_method, dist_method){
# ordinate
ps_ordination <- ordinate(filtered_ps, method = vis_method, distance = dist_method)
# get eignenvalues
evals <- ps_ordination$values$Eigenvalues
# generate plot
plot_ordination(filtered_ps, ps_ordination, color = variable, shape = variable_2,
  title = ) +
  labs(col = variable) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  geom_point(aes(label = ID), size = 5, alpha = 0.5) +
  geom_text(aes(label = ifelse(ID == 319, as.character(Sample_Type), '')), hjust = 0, vjust = 2) +
  scale_colour_viridis_d()
 }

ordination_plots(ps2_storage, "URN", "Sample_Type", "PCoA", "bray") +
  scale_shape_discrete(name = "Storage") 

ggsave("outputs/PCoA.png", dpi = 600, height = 3, width = 5)
```

**319 = OMNIgene sample, not original sample. However, it does have the greatest read depth of the samples within this URN. Thus, I hypothesise that the greater read depth means that more diversity was captured by this particular sample, and that although we have filtered out those taxa not frequently observed and normalised the data with TSS, these methods (no methods) are perfect. For instance, sample 319 may have some taxa that are in low abundance within the context of URN 648890, but are still widely distributed throughout the other URNs, and thus are not removed and can still contribute to a greater distance to its technical replicates.**


### Statistical test: PERMANOVA.
 - ps2 transformed.
 - Preforming permutational anova for group-level differences based on dissimilarity.
 - Extract otu table and metadata from phyloseq object.
 - Use `adonis()` from the *vegan* package to perform the PERMANOVA.
 - Homogeneity Condition
  - Significant PERMANOVA means one of three things:
  - there is a difference in the location of the samples (i.e. the average community composition).
  - there is a difference in the dispersion of the samples (i.e. the variability in the community composition).
  - there is a difference in both the location and the dispersion.
  - If you get a significant PERMANOVA you'll want to distinguish between the three options by checking the homogeneity condition using `permdisp()`. If you get a non-significant result the first option above is correct.
  - `betadisper()` gives a measure of the dispersion within groups. Thus, if the PERMANOVA test is significant and the permdisp is not, the significant result in your communities is due to a mean shift in community composition and not from increased variance within groups.
  
#### Mixed effects comparison with technical replicates nested within individuals

#### Mixed effects comparison of storage conditions and URN as main effects

```{r, warning=F, message=F}
# permanova
ps_otu <- data.frame(otu_table(ps2_storage))
ps_metadata <- data.frame(sample_data(ps2_storage))
permanova <- adonis2(ps_otu ~Sample_Type + URN, data = ps_metadata, method = "bray", by = "margin") # adnois 2
permanova 
```

##### Export
```{r, warning=F, message=F, eval=F}
permanova_func <- function(ps2_storage){
  
# permanova
ps_otu <- data.frame(otu_table(ps2_storage))
ps_metadata <- data.frame(sample_data(ps2_storage))
permanova <- adonis2(ps_otu ~Sample_Type + URN, data = ps_metadata, method = "bray", by = "margin") # adnois 2
permanova <- tableGrob(as.data.frame(permanova)) %>% 
  annotate_figure(fig.lab = "PERMANOVA", fig.lab.face = "bold", fig.lab.size = 15)

# homogeneity condition
dist <- vegdist(ps_otu)
homogeneity <- as.data.frame(anova(betadisper(dist, ps_metadata$Sample_Type))) %>% 
  tableGrob() %>% 
  annotate_figure(fig.lab = "Homogeneity Condition", fig.lab.face = "bold", fig.lab.size = 15)

# combine ouputs in a grid
grid.arrange(permanova, homogeneity, ncol = 1)

}

permanova_func(ps2_storage)
```

#### Mixed effects comparison with technical replicates nested within individuals using envfit

```{r,warning=F,message=F,eval=F}
ordination_distance <- ordinate(
    physeq = ps2_storage, 
    method = "NMDS",
    distance = "bray",
    formula = ~ Sample_Type + URN)

envdat.phy <- ps2_storage %>% 
  sample_data() %>%
  unclass() %>% 
  as.data.frame() %>% 
  select(Sample_Type, URN) 

set.seed(1)

envfit_object <- envfit(ordination_distance, envdat.phy, permutations = 10000) 
``` 

# Adjust p values

```{r,warning=F,message=F,eval=F}
p.adjust.envfit <- function (x, method = 'bonferroni', n){
  x.new <- x
  if (!is.null (x$vectors)) pval.vectors <- x$vectors$pvals else pval.vectors <- NULL
  if (!is.null (x$factors)) pval.factors <- x$factors$pvals else pval.factors <- NULL
  if (missing (n)) n <- length (pval.vectors) + length (pval.factors)
  if (!is.null (x$vectors)) x.new$vectors$pvals <- p.adjust (x$vectors$pvals, 
                                                             method = method, n = n)
  if (!is.null (x$factors)) x.new$factors$pvals <- p.adjust (x$factors$pvals, 
                                                             method = method, n = n)
  cat ('Adjustment of significance by', method, 'method')
  return (x.new)
}

set.seed(1)

envfit(ordination_distance, envdat.phy, permutations = 10000) %>% 
  p.adjust.envfit(., method = "BH") 
```

# Plot model

```{r,warning=F,message=F}
envfit_object$factors$r %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  dplyr::rename("r2" = ".") %>%  # changed to 0.1 for plot 
  left_join(
    envfit_object$factors$pvals %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  dplyr::rename("p" = ".")
  ) %>% 
  ggplot(aes(x = r2, y = Variable, fill = p)) +
  geom_col() +
  geom_point(size = 5) +
  theme(axis.line =  element_line(size = .5), panel.background = element_blank()) + 
  scale_fill_continuous(type = "viridis") +
  scale_y_discrete(labels = c("Sample_Type" = "Storage")) +
  ylab("Covariates")
```

#### EXport plots

```{r, warning=F, message=F, eval=F}
ordination_plots <- function(filtered_ps, variable, vis_method, dist_method){
# ordinate
ps_ordination <- ordinate(filtered_ps, method = vis_method, distance = dist_method)
# get eignenvalues
evals <- ps_ordination$values$Eigenvalues
# generate plot
plot_ordination(filtered_ps, ps_ordination, color = variable, 
  title = "") +
  labs(col = variable) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  geom_point(aes(label = ID), size = 5, alpha = 0.5) +
  geom_text(aes(label = ifelse(ID == 319, as.character(Sample_Type), '')), hjust = 0, vjust = 2) +
  scale_colour_viridis_d()
}

PCoA <- ordination_plots(ps2_storage, "URN", "PCoA", "bray")

envfit <- envfit_object$factors$r %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  dplyr::rename("r2" = ".") %>%  # changed to 0.1 for plot 
  left_join(
    envfit_object$factors$pvals %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  dplyr::rename("p" = ".")
  ) %>% 
  ggplot(aes(x = r2, y = Variable, fill = p)) +
  geom_col() +
  geom_point(size = 5) +
  theme(axis.line =  element_line(size = .5), panel.background = element_blank(), aspect.ratio = 4/3) + 
  scale_fill_continuous(type = "viridis") +
  scale_y_discrete(labels = c("Sample_Type" = "Storage")) +
  ylab("Covariates") 
  
grid_plot <- grid.arrange(PCoA, envfit, ncol=2)

ggsave("outputs/PCoA.png", dpi = 600, height = 6, width = 6, PCoA)

ggsave("outputs/envfit.png", dpi = 600, height = 5, width = 5, envfit)

ggsave("outputs/grid_plot.png", dpi = 600, height = 5, width = 7, grid_plot)
```


## Taxanomic abundance.

### Bar charts
 - Use `plot_bar_auto()` function wrapped around phyloseq's `plot_bar()` to explore the distribution of taxa at the genus and phylum levels.
 - Subset transformed data (relative abundance) to only the top20 taxa.
 
```{r, warning=F, message=F, fig.cap="Bar plots of the taxonomic distribution (relative abundance) at the genus levels."}
top20 <- names(sort(taxa_sums(ps4_storage), decreasing=TRUE))[1:20]
ps.top20 <- prune_taxa(top20, ps4_storage)


sample_data(ps.top20) <- sample_data(ps.top20) %>% 
  unclass() %>% 
  as.data.frame() %>% 
  mutate(URN = paste0("URN ", URN)) %>% 
  mutate("Sample" = ID) %>% 
  mutate(OMNIgene_Number = ifelse(OMNIgene_Number == "Original", "Original", "OMNIgene")) %>% 
  column_to_rownames("Sample") 

modified_plot_bar <- function (physeq, x = "Sample", y = "Abundance", 
    fill = NULL, title = NULL, facet_grid = NULL, alpha = NULL){
    mdf = psmelt(physeq)
    p = ggplot(mdf, aes_string(x = x, y = y, fill = fill, alpha = alpha))
    p = p + geom_bar(stat = "identity", position = "stack", 
        color = "black")
    p = p + theme(axis.text.x = element_text(angle = -90, hjust = 0))
    if (!is.null(facet_grid)) {
        p <- p + facet_grid(facet_grid)
    }
    if (!is.null(title)) {
        p <- p + ggtitle(title)
    }
    return(p)
}

plot_bar_auto <- function(ps, taxonomy, alpha){
modified_plot_bar(ps, fill = taxonomy, alpha = alpha) + 
    facet_wrap(~URN, scales = "free_x") + 
    labs(title = paste0("Level:", taxonomy), y = "Abundance") + 
    theme(legend.position = "bottom", legend.title = element_blank(),
    axis.title.x = element_blank(), axis.text.x = element_blank(), 
    axis.ticks = element_blank()) +
    scale_fill_viridis_d() +
    scale_alpha_discrete(range = c(1, 0.5), name = "Storage") 
}

grid.arrange(plot_bar_auto(ps.top20, "Phylum", "OMNIgene_Number"),
             plot_bar_auto(ps.top20, "Genus", "OMNIgene_Number"),
             nrow = 2, heights = c(1, 1))
```


#### Export Plot
```{r,warning=F,message=F,eval=F}
grid_plot <- grid.arrange(plot_bar_auto(ps.top20, "Phylum", "OMNIgene_Number"),
             plot_bar_auto(ps.top20, "Genus", "OMNIgene_Number"),
             nrow = 2, heights = c(1, 1))

ggsave("outputs/taxonomy_plot.png", dpi = 600, height = 10, width = 10, grid_plot)
```

## Perform DESeq2 analysis

 - Define function for calculating geometric means and estimating size factors.
 - Define function to filter out taxa with small counts and low occurance. *count* and *samples* arguments need to be applied as numerical values.
 
```{r, warning=F, message=F}
calc_geo_means <- function(deseq_object){
# geometric mean
  gm_mean = function(x, na.rm = TRUE){
    exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
  }
  geoMeans <- apply(counts(deseq_object), 1, gm_mean)
# size factors
  estimateSizeFactors(deseq_object, geoMeans = geoMeans) 
}

deseq_filter <- function(deseq_object, count, samples){ 
  nc <- counts(deseq_object, normalized = TRUE)
  filtered <- rowSums(nc >= count) >= samples 
  deseq_object[filtered,]
}
```

 - Define a function to extract the results.
 - Convert from *phyloseq* to *deseq* object.
 - Use prevviously defined functions to calculate geometric means and filter to the most abundant and frequent taxa.
 - Use `Deseq()` to perform the normalisation and analysis.
 - Extract the results using appropriate previosuly defined function.
 
```{r, warning=F,message=F}
# define function for Wald test
get_deseq_res_cat <- function(desq_object, contrast_variable, level1, level2){
  res = results(desq_object, contrast = c(contrast_variable, level1, level2))
  res = res[order(res$padj, na.last = NA), ]
  sigtab = res[(res$padj < 0.05), ] 
  sigtab = cbind(as(sigtab, "data.frame"), 
    as(tax_table(ps3_storage)[rownames(sigtab), ], "matrix"))
  sigtab %>%
  arrange(padj) %>%
  select("log2FoldChange", "lfcSE", "padj", "Genus") %>%
  add_column(Variable = paste0(contrast_variable, level1)) # label the base level
}

phyloseq_to_deseq2(ps3_storage, ~ Sample_Type + URN) %>% 
                  calc_geo_means() %>% 
                  deseq_filter(1, 5) %>% # count > 1 in at 5 samples
                  DESeq(fitType = "local", test = "Wald") %>% 
                  get_deseq_res_cat("Sample_Type", "Original", "OMNIgene") %>% 
                  remove_rownames() %>% 
                  knitr::kable()
```

## Alpha diversity.
 - Define a function that calculates Shannon Index, Obsverved (richness) & Chao1 diversity, and binds it to our original metadata dataframe, which can then be used for analysis.
 
```{r,warning=F,message=F}
ps2_storage <- subset_samples(ps2, Study == "2") %>% 
  subset_samples(URN != "957652") %>% 
  subset_samples(URN != "648892")
```


```{r, warning=F, message=F, eval=F}
calc_alpha_diversity <- function(ps2_storage){
# calculate metrics
ps_alpha_div <- ps2_storage %>%
                estimate_richness(measures = c("Shannon", "Observed", "Chao1")) %>% 
                select(-se.chao1)

# creat ID column based on rownames
ps_alpha_div <- rownames_to_column(ps_alpha_div, var = "ID") %>% 
                mutate(ID = as.factor(gsub("X", "", ID)))

# join alpha metrics with metadata by the ID column
Metadata %>%
  filter(Study == "2") %>% 
  right_join(ps_alpha_div, by = "ID") %>%
  as.data.frame() 
}

ps_metadata <- calc_alpha_diversity(ps2_storage)
```

### Distribution

#### Histogram
 
```{r, warning=F, message=F}
# To determine if diveristy is normally distributed
ggplot(ps_metadata, aes(x = Shannon)) + 
  geom_histogram() + 
  xlab("Alpha Diversity") + ylab("Count")
```

#### Test for normality.
 
```{r, warning=F, message=F}
shapiro.test(ps_metadata$Shannon)
```

### Plot 

#### Point

 - Use `plot_richness()` from *phyloseq*, which estimates alpha diversity metrics using *vegan* and plots them, taking standard *ggplot2* *geoms_* for the plot design.
 - use ps2 non-transformed data for alpha.
 
```{r, warning=F, message=F}
plot_richness(ps2_storage, measures = c("Shannon", "Observed"), 
              color = "URN", title = "", shape = "Sample_Type") +
    geom_point(size = 3.5, alpha = 0.7) +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          panel.border = element_rect(colour = "grey", fill = NA, size = 1))+
  scale_color_viridis_d() +
  xlab("Technical Replicates") +
  ylab("Alpha Diversity") +
  scale_shape_discrete("Storage")
```

#### Export plot
```{r, warning=F, message=F}
ggsave("outputs/alpha_plot.png", dpi = 600 , height = 8, width = 8)
```

#### Boxplot comparing storage types
```{r, warning=F, message=F}
ps_metadata %>% 
  ggplot(aes(x = Sample_Type, y = Shannon, colour = Sample_Type)) +
  geom_boxplot() +
  labs(x = "Storage Method", y = "Shannon Diversity") +
  stat_compare_means() +
  theme(legend.position = "none") +
  scale_color_viridis_d()
```

#### Comparing storage types within URNs

```{r}
plot_richness(ps2_storage, measures = c("Shannon"), 
              color = "Sample_Type", title = "") +
    geom_point(size = 3.5, alpha = 0.7) +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          panel.border = element_rect(colour = "grey", fill = NA, size = 1))+
  scale_color_viridis_d() + 
  facet_wrap(~URN) 
```

### Statistical test: compare mean/median between groups.
- define a function that performs a Wilcoxin test on the three diversity metrics and binds them (Shannon Index, Richness & Chao1).

```{r, warning=F, message=F}
diversity_analysis <- function(ps_metadata){

Shannon <- compare_means(Shannon ~ Sample_Type, data = ps_metadata, 
                         method = "wilcox.test", p.adjust.method = "fdr")

Observed <- compare_means(Observed ~ Sample_Type, data = ps_metadata, 
                          method = "wilcox.test", p.adjust.method = "fdr")

Chao1 <- compare_means(Chao1 ~ Sample_Type, data = ps_metadata, 
                       method = "wilcox.test", p.adjust.method = "fdr")

bind_rows(Shannon, Observed, Chao1) %>%
  rename(c(".y." = "Diversity Measure"))
}

diversity_analysis(ps_metadata)
```

### GLM 

**Based on same formula as PERMANOVA**

### Test for collinearity using known microbiome-covariates

```{r,warning=F,message=F,eval=F}
# defin myvif function
myvif <- function(mod) {
  v <- vcov(mod)
  assign <- attributes(model.matrix(mod))$assign
  if (names(coefficients(mod)[1]) == "(Intercept)") {
    v <- v[-1, -1]
    assign <- assign[-1]
  } else warning("No intercept: vifs may not be sensible.")
  terms <- labels(terms(mod))
  n.terms <- length(terms)
  if (n.terms < 2) stop("The model contains fewer than 2 terms")
  if (length(assign) > dim(v)[1] ) {
    diag(tmp_cor)<-0
    if (any(tmp_cor==1.0)){
      return("Sample size is too small, 100% collinearity is present")
    } else {
      return("Sample size is too small")
    }
  }
  R <- cov2cor(v)
  detR <- det(R)
  result <- matrix(0, n.terms, 3)
  rownames(result) <- terms
  colnames(result) <- c("GVIF", "Df", "GVIF^(1/2Df)")
  for (term in 1:n.terms) {
    subs <- which(assign == term)
    result[term, 1] <- det(as.matrix(R[subs, subs])) * det(as.matrix(R[-subs, -subs]))/detR
    result[term, 2] <- length(subs)
  }
  if (all(result[, 2] == 1)) {
    result <- data.frame(GVIF=result[, 1])
  } else {
    result[, 3] <- result[, 1]^(1/(2 * result[, 2]))
  }
  invisible(result)
}

# corvif
corvif <- function(data) {
  data <- as.data.frame(data)

  form    <- formula(paste("fooy ~ ",paste(strsplit(names(data)," "),collapse = " + ")))
  data  <- data.frame(fooy = 1 + rnorm(nrow(data)) ,data)
  lm_mod  <- lm(form,data) # runs linear model with above formula and metadata
  
  cat("\n\nVariance inflation factors\n\n")
  print(myvif(lm_mod))
}

ps_metadata %>% 
  select(URN, Sample_Type) %>% 
  corvif()

```

### Fit Model - Shannon

```{r,warning=F,message=F}
global <- glm(Shannon ~ Sample_Type + URN, data = ps_metadata, 
                        family = Gamma(link = "log")) 

global

car::Anova(global) %>% 
  as.data.frame(row.names = NULL)

emmeans(global, list(pairwise ~ Sample_Type), adjust = "tukey") %>% 
  pairs() %>% 
  as.data.frame() 
```

```{r,warning=F,message=F,eval=F}
gof(global)
r.squaredGLMM(global)
```

### Fit Model - Richness

```{r,warning=F,message=F}
global <- glm(Observed ~ Sample_Type + URN, data = ps_metadata, 
                        family = Gamma(link = "log")) 

global

car::Anova(global) %>% 
  as.data.frame(row.names = NULL)

emmeans(global, list(pairwise ~ Sample_Type), adjust = "tukey") %>% 
  pairs() %>% 
  as.data.frame() 
```

```{r,warning=F,message=F,eval=F}
gof(global)
r.squaredGLMM(global)
```


# Read depth

```{r,warning=F,message=F,eval=F}
ps_storage <- subset_samples(ps, Study == "2") %>% 
  subset_samples(URN != "957652") %>% 
  subset_samples(URN != "648892")

read_depth <- sample_data(ps_storage) %>% 
  unclass() %>% 
  as.data.frame() %>% 
  mutate(TotalReads = as.numeric(sample_sums(ps_storage))) %>%
  arrange(URN) %>% 
  add_column(ID_Plot = 1:nrow(.)) 
  
read_depth %>% 
  mutate(OMNIgene_Number = ifelse(OMNIgene_Number == "Original", "Original", "OMNIgene")) %>% 
  ggplot(aes(y = TotalReads, x = ID_Plot, fill = URN, alpha = OMNIgene_Number)) + 
  geom_col() +
  ggtitle("") +
  xlab("Technical Replicates") +
  ylab("Read depth")  +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  scale_alpha_discrete(range = c(1, 0.5), name = "Storage") 
```

#### Export plot
```{r, warning=F, message=F}
ggsave("outputs/read_depth.png", dpi = 600 , height = 6, width = 9)
```

#### Test for normality.
 
```{r, warning=F, message=F}
shapiro.test(read_depth$TotalReads)
```

```{r,warning=F,message=F}
global <- glm(TotalReads ~ Sample_Type + URN, data = read_depth)  

global

car::Anova(global) %>% 
  as.data.frame(row.names = NULL)

emmeans(global, list(pairwise ~ Sample_Type), adjust = "tukey") %>% 
  pairs() %>% 
  as.data.frame() 
```



