---
title: "Shotgun Analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/Jacob/Desktop/Jacob_Uni/Data/Long_term_effects_of_probiotics')
```

# Exploratory analysis with shotgun metagenomics data

Sequencing and pre-analytical bioinformatics performed by Microba

## About.


## Load required packages. 

```{r, warning=F, message=F, results='hide'}
sapply(c("caret", "pls", "e1071", "ggplot2", 
    "randomForest", "tidyverse", "ggrepel", "nlme", "devtools", 
    "reshape2", "PMA", "structSSI", "ade4","ggnetwork", 
    "intergraph", "scales", "readxl", "genefilter", "impute", 
    "phyloseq", "phangorn", "dada2", "DECIPHER", "gridExtra", "stringi", "janitor"), 
    require, character.only = TRUE)
```

## Import taxonomy

```{r, warning=F, message=F, eval=F}
shotgun_taxa <- read_tsv("data/shotgun/profiles.tsv") %>% 
  select(1:2) %>% 
  separate(2, sep = ";", c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")) %>% 
  dplyr::rename("Species" = "Taxon") %>% 
  relocate(c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")) %>% 
  mutate_at(c(1:ncol(.)), ~str_remove(., ".*__")) %>% 
  mutate(ASV = Species) %>% 
  mutate(ASV = str_replace(ASV, " ", "_")) %>% 
  column_to_rownames("ASV") %>% 
  as.matrix()
```

## Import taxonomy

```{r, warning=F, message=F, eval=F}
shotgun_ASV <- read_tsv("data/shotgun/profiles.tsv") %>% 
  select(-2) %>% 
  mutate(Taxon = str_remove(Taxon, ".*__")) %>% 
  mutate(Taxon = str_replace(Taxon, " ", "_")) %>% 
  column_to_rownames("Taxon") %>% 
  t() %>% 
  as.matrix()
```

## Import metadata

```{r, warning=F, message=F, eval=F}
shotgun_metadata <- readxl::read_excel("data/shotgun/Metadata.xlsx") %>% 
  select(1:2, 10:ncol(.)) %>% 
  dplyr::rename("ID" = "Microba ID") %>% 
  as.data.frame() %>% 
  mutate(sam_names = ID) %>% 
  column_to_rownames("sam_names") %>% 
  dplyr::rename("Client_ID" = "Client ID",
         "Probiotics" = "Probitoics during admission" ,
         "Delivery" = "Mode of Delivery",
         "Diet" = "Feeding Type (during admission)",
         "Formula" = "Ever given formula") %>% 
  mutate_if(is.numeric, as.factor) %>% 
  mutate(Probiotics = ifelse(Probiotics == "0", "No", "Yes")) %>% 
  mutate(Gestation = str_remove(Gestation, " Weeks")) %>% 
  separate(Gestation, sep = "\\+", into = c("Weeks", "Days")) %>% 
  mutate_at(c("Weeks", "Days"), as.numeric) %>% 
  mutate(Days = ifelse(is.na(Days), 0, .$Days)) %>% 
  mutate(Gestation = (Weeks*7)+Days)
```

## Constrcut the Phyloseq object.
 - Includes: metadata, ASV table, taxonomy table and phylogenetic tree.
 
```{r, warning=F, message=F, results='hide', eval=F}
ps <- phyloseq(otu_table(shotgun_ASV, taxa_are_rows=FALSE), 
               sample_data(shotgun_metadata), 
               tax_table(shotgun_taxa))
```


## Getting read counts
```{r, eval=F}
sample_data(ps) %>% 
  unclass() %>% 
  as.data.frame() %>% 
  mutate(TotalReads = sample_sums(ps)) %>% 
  ggplot(aes(TotalReads)) + 
    geom_histogram() + 
    ggtitle("Sequencing Depth")
```

## Filtering and normalisation.

### Taxonomy filtering.
 - Can check the number of phyla before and after transformation with `table(tax_table(ps)[, "Phylum"], exclude = NULL)`.
 - Remove features with ambiguous and NA phylum annotation.

```{r, warning=F, message=F, results='hide', eval=F}
ps1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
```

### Filtering.
 - Using an unsupervised method (relying on the data in this experiment) explore the prevelance of features in the dataset.
 - Calculate the prevalence of each feature and store as a dataframe.
 - Add taxonomy and total read counts.
 
```{r, warning=F, message=F, results='hide', eval=F}
prevdf = apply(X = otu_table(ps1),
               MARGIN = ifelse(taxa_are_rows(ps1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps1),
                    tax_table(ps1))
```

 - Plot the relationship between prevelance and total read count for each feature. This provides information on outliers and ranges of features.
 
```{r, warning=F, message=F, fig.cap="Scatterplot exploring the relationship between prevelance and abundance of phyla."}
prevdf %>%
  subset(Phylum %in% get_taxa_unique(ps1, "Phylum")) %>%
  ggplot(aes(TotalAbundance, Prevalence / nsamples(ps1),color=Phylum)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 1) +  
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  
  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```
 
 - Define abundance threshold
 
```{r, warning=F, message=F, results='hide', eval=F}
prevalenceThreshold = prevdf %>% 
  summarise(sum(TotalAbundance)*.00001) %>% 
  pull(1)

keepTaxa = prevdf %>% 
  filter(TotalAbundance >= prevalenceThreshold) %>% 
  rownames()

ps2 = prune_taxa(keepTaxa, ps1)

as.data.frame(keepTaxa) %>%  nrow()
```

**Used abundance threshold.**

 - Explore the relationship on the filtered data set.
```{r, warning=F, message=F, fig.cap="Scatterplot exploring the relationship between prevelance and abundance of phyla on data passed through a prevalence threshold."}
prevdf %>%
  subset(Phylum %in% get_taxa_unique(ps2, "Phylum")) %>%
  ggplot(aes(TotalAbundance, Prevalence / nsamples(ps2),color=Phylum)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 1) +  
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  
  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

#### Check percentages of NA values left.
```{r,warning=F,message=F}
# Per taxonomic rank
apply(tax_table(ps2), 2, function(col)sum(is.na(col))/length(col)) * 100
```

#### Check percentages of NA values left.
```{r,warning=F,message=F}
# Per taxonomic rank
apply(tax_table(ps3), 2, function(col)sum(is.na(col))/length(col)) * 100 
```

### Normalisation.
 - Plot a refraction curve to see if total sum scaling will surfice.
 - Define colours and lines.
 - Step = step size for sample sizes in rarefaction curve.
 
```{r, warning=F, message=F, eval=F}
vegan::rarecurve(t(otu_table(ps3)), step = 20, label = FALSE, main = "Rarefaction Curve", 
        col = c("black", "darkred", "forestgreen", "orange", "blue", "yellow", "hotpink"))
```

 - Perform total sum scaling on agglomerated dataset.
 
```{r, warning=F, message=F, eval=F}
ps4 <- transform_sample_counts(ps2, function(x) x / sum(x))
```


# Data exploration and univariate analysis.


## About

This script explores the differences and similarities between sample storage types, and uses ordinations and EnvFit/PERMANOVA analysis (Vegan package) to determine if storage of faecal samples with the OMNIGene collection kit differed from the gold standard freezing at -80. The EnvFit function was applied to an ordination using NMDS that was based on a Bray-Curtis dissimilarity matrix calculated from data normalised with Total Sum Scaling, and the PERMANOVA to an ordination using PCoA that was also based on Bray-Curtis. The significance of both was based on 10,000 permutations and was transformed based on the Benjamini-Hochberg (BH) procedure.

## Load required packages. 

```{r, warning=F, message=F, results='hide'}
sapply(c("BiocManager", "ggplot2", "ggforce", "vegan", "knitr", "dplyr", 
         "phyloseq", "phyloseqGraphTest", "igraph", "ggnetwork", "nlme", 
         "reshape2", "tidyverse", "plyr", "DESeq2", "sjPlot", "ggpubr", 
         "gridExtra", "grid", "gtable", "lazyeval", "phyloseq", "tidyverse", 
         "knitr", "lme4", "emmeans", "MuMIn", "aods3", "sjPlot", "car"), 
       require, character.only = TRUE)
```

## Beta diversity
 - Use distance and ordination methods to explore the relationship between metadata.
 - We calculate the distances using pruned, transformed (TSS) and non-agglomerated data.

```{r, warning=F, message=F, eval=F}
ps2 <- ps2 %>%
        transform_sample_counts(function(x) x / sum(x))
```
 
 - We can then create distance matrices and plots for this data subset using several methods:
 - e.g. bray-curtis or weighted unifrac distances with PCoA, NMDS, etc.
 - Define a function that ordindates the previously transformed data, extarcts the eigenvalues, and creates a dissimilarity plot.
 - Extract eigenvalues from ordination.

```{r, fig.cap="PCoA plot of Bray-Curtis distances coloured by date.", warning=F, message=F}
ordination_plots <- function(filtered_ps, variable, vis_method, dist_method){
# ordinate
ps_ordination <- ordinate(filtered_ps, method = vis_method, distance = dist_method)
# get eignenvalues
evals <- ps_ordination$values$Eigenvalues
# generate plot
plot_ordination(filtered_ps, ps_ordination, color = variable, 
  title = "") +
  labs(col = variable) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  geom_point(aes(label = ID), size = 5) +
  geom_text(aes(label = ifelse(ID == 319, as.character(Sample_Type), '')), hjust = 0, vjust = 2) +
  scale_color_manual(values = c("#440154FF", "#55C667FF"))
 }

ordination_plots(ps2, "Probiotics", "PCoA", "bray")

#ggsave("outputs/PCoA.png", dpi = 600, height = 5, width = 7)
```

#### Export
```{r}
PCoA <- ordination_plots(ps2, "Probiotics", "PCoA", "bray")
ggsave("outputs/shotgun/PCoA.png", dpi = 600, height = 5, width = 6, PCoA)
```

### Statistical test: PERMANOVA.
 - ps2 transformed.
 - Preforming permutational anova for group-level differences based on dissimilarity.
 - Extract otu table and metadata from phyloseq object.
 - Use `adonis()` from the *vegan* package to perform the PERMANOVA.
 - Homogeneity Condition
  - Significant PERMANOVA means one of three things:
  - there is a difference in the location of the samples (i.e. the average community composition).
  - there is a difference in the dispersion of the samples (i.e. the variability in the community composition).
  - there is a difference in both the location and the dispersion.
  - If you get a significant PERMANOVA you'll want to distinguish between the three options by checking the homogeneity condition using `permdisp()`. If you get a non-significant result the first option above is correct.
  - `betadisper()` gives a measure of the dispersion within groups. Thus, if the PERMANOVA test is significant and the permdisp is not, the significant result in your communities is due to a mean shift in community composition and not from increased variance within groups.


```{r, warning=F, message=F}
# permanova
ps_otu <- data.frame(otu_table(ps2))
ps_metadata <- data.frame(sample_data(ps2))
permanova <- adonis2(ps_otu ~Probiotics + Delivery + Diet, data = ps_metadata, method = "bray", by = "margin") # adnois 2
permanova 
```

#### Export
```{r, warning=F, message=F, eval=F}
permanova_func <- function(ps2){
  
# permanova
ps_otu <- data.frame(otu_table(ps2))
ps_metadata <- data.frame(sample_data(ps2))
permanova <- adonis2(ps_otu ~Probiotics + Delivery + Diet, data = ps_metadata, method = "bray", by = "margin") # adnois 2
permanova <- permanova %>% 
  as.data.frame() %>% 
  mutate_at(c(2:5), ~round(., digits = 2)) %>% 
  tableGrob() %>% 
  annotate_figure(fig.lab = "PERMANOVA", fig.lab.face = "bold", fig.lab.size = 15)

# homogeneity condition
dist <- vegdist(ps_otu)
homogeneity <- as.data.frame(anova(betadisper(dist, ps_metadata$Probiotics))) %>% 
  mutate_at(c(2:5), ~round(., digits = 2)) %>% 
  tableGrob() %>% 
  annotate_figure(fig.lab = "Homogeneity Condition", fig.lab.face = "bold", fig.lab.size = 15)

# combine ouputs in a grid
grid.arrange(permanova, homogeneity, ncol = 1)

}

permanova_table <- permanova_func(ps2)

ggsave("outputs/shotgun/permanova.png", dpi = 600, height = 5, width = 5, permanova_table)
```

#### Statisitcal test: envfit

```{r,warning=F,message=F,eval=F}
ordination_distance <- ordinate(
    physeq = ps2, 
    method = "NMDS",
    distance = "bray",
    formula = ~Probiotics + Delivery + Diet)

envdat.phy <- ps2 %>% 
  sample_data() %>%
  unclass() %>% 
  as.data.frame() %>% 
  select(Probiotics, Delivery, Diet) 

set.seed(1)

envfit_object <- envfit(ordination_distance, envdat.phy, permutations = 10000) 
``` 

# Adjust p values

```{r,warning=F,message=F,eval=F}
p.adjust.envfit <- function (x, method = 'bonferroni', n){
  x.new <- x
  if (!is.null (x$vectors)) pval.vectors <- x$vectors$pvals else pval.vectors <- NULL
  if (!is.null (x$factors)) pval.factors <- x$factors$pvals else pval.factors <- NULL
  if (missing (n)) n <- length (pval.vectors) + length (pval.factors)
  if (!is.null (x$vectors)) x.new$vectors$pvals <- p.adjust (x$vectors$pvals, 
                                                             method = method, n = n)
  if (!is.null (x$factors)) x.new$factors$pvals <- p.adjust (x$factors$pvals, 
                                                             method = method, n = n)
  cat ('Adjustment of significance by', method, 'method')
  return (x.new)
}

set.seed(1)

envfit_object <- envfit(ordination_distance, envdat.phy, permutations = 10000) %>% 
  p.adjust.envfit(., method = "BH")

envfit_object$factors %>% 
  unclass() %>% 
  as.data.frame()
```

# Differential abundance

## Taxanomic distribution.

### Bar charts
 - Use `plot_bar_auto()` function wrapped around phyloseq's `plot_bar()` to explore the distribution of taxa at the genus and phylum levels.
 - Subset transformed data (relative abundance) to only the top20 taxa.
 
```{r, warning=F, message=F, fig.cap="Bar plots of the taxonomic distribution (relative abundance) at the genus levels."}
top20 <- names(sort(taxa_sums(ps4), decreasing=TRUE))[1:30]
ps.top20 <- prune_taxa(top20, ps4)

plot_bar_auto <- function(ps, taxonomy){
plot_bar(ps, fill = taxonomy) + 
    geom_bar(aes(fill=Genus), stat="identity", position="stack") +
    facet_wrap(~Probiotics, scales = "free_x") + 
    labs(title = paste0("Level:", taxonomy), y = "Abundance") + 
    theme(legend.position = "bottom", legend.title = element_blank(), 
    axis.title.x = element_blank(), axis.text.x = element_blank(), 
    axis.ticks = element_blank()) +
    scale_fill_viridis_d()
}

grid.arrange(plot_bar_auto(ps.top20, "Phylum"),
             plot_bar_auto(ps.top20, "Species"),
             nrow = 2, heights = c(1, 1.2))
```

#### Export Plot
```{r,warning=F,message=F,eval=F}
plot_bar_auto(ps.top20, "Species")

ggsave("outputs/shotgun/taxonomy_plot.png", dpi = 600, height = 6, width = 12)
```

```{r, warning=F, message=F}
top20 <- names(sort(taxa_sums(ps4), decreasing=TRUE))[1:50]
ps.top20 <- prune_taxa(top20, ps4)

plot_bar_auto <- function(ps, taxonomy){
plot_bar(ps, fill = taxonomy) + 
    facet_wrap(~Probiotics, scales = "free_x") + 
    labs(title = paste0("Level:", taxonomy), y = "Abundance") + 
    theme(legend.position = "none", legend.title = element_blank(), 
    axis.title.x = element_blank(), axis.text.x = element_blank(), 
    axis.ticks = element_blank()) +
    scale_fill_viridis_d()
}

grid.arrange(plot_bar_auto(ps.top20, "Phylum"),
             plot_bar_auto(ps.top20, "Species"),
             nrow = 2, heights = c(1, 1))
```



## Perform DESeq2 analysis

```{r, warning=F, message=F}
ps2 = prune_taxa(keepTaxa, ps1)
```

 - Define function for calculating geometric means and estimating size factors.
 - Define function to filter out taxa with small counts and low occurance. *count* and *samples* arguments need to be applied as numerical values.
 
```{r, warning=F, message=F}
calc_geo_means <- function(deseq_object){
# geometric mean
  gm_mean = function(x, na.rm = TRUE){
    exp(sum(log(x[x > 0]), na.rm = na.rm) / length(x))
  }
  geoMeans <- apply(counts(deseq_object), 1, gm_mean)
# size factors
  estimateSizeFactors(deseq_object, geoMeans = geoMeans) 
}

deseq_filter <- function(deseq_object, count, samples){ 
  nc <- counts(deseq_object, normalized = TRUE)
  filtered <- rowSums(nc >= count) >= samples 
  deseq_object[filtered,]
}
```

 - Define a function to extract the results.
 - Convert from *phyloseq* to *deseq* object.
 - Use prevviously defined functions to calculate geometric means and filter to the most abundant and frequent taxa.
 - Use `Deseq()` to perform the normalisation and analysis.
 - Extract the results using appropriate previosuly defined function.
 
```{r, warning=F,message=F}
# define function for Wald test
get_deseq_res_cat <- function(desq_object, contrast_variable, level1, level2){
  res = results(desq_object, contrast = c(contrast_variable, level1, level2))
  res = res[order(res$padj, na.last = NA), ]
  sigtab = res[(res$padj < 0.05), ] 
  sigtab = cbind(as(sigtab, "data.frame"), 
    as(tax_table(ps2)[rownames(sigtab), ], "matrix"))
  sigtab %>%
  arrange(padj) %>%
  select("log2FoldChange", "lfcSE", "padj", "Species") %>%
  add_column(Variable = paste0(contrast_variable, level1)) # label the base level
}

phyloseq_to_deseq2(ps2, ~Probiotics + Delivery + Diet) %>% 
                  calc_geo_means() %>% 
                  deseq_filter(10, 10) %>% 
                  DESeq(fitType = "local", test = "Wald") %>% 
                  get_deseq_res_cat("Probiotics", "Yes", "No") %>% 
                  remove_rownames() %>% 
                  knitr::kable()
```


## Alpha diversity.
 
```{r,warning=F,message=F}
ps2 = prune_taxa(keepTaxa, ps1)
```


```{r, warning=F, message=F, eval=F}
calc_alpha_diversity <- function(ps2){
# calculate metrics
ps_alpha_div <- ps2 %>%
                estimate_richness(measures = c("Shannon", "Observed", "Chao1")) %>% 
                select(-se.chao1)

# creat ID column based on rownames
ps_alpha_div <- rownames_to_column(ps_alpha_div, var = "ID") %>% 
                mutate(ID = as.factor(gsub("X", "", ID)))

# join alpha metrics with metadata by the ID column
shotgun_metadata %>%
  right_join(ps_alpha_div, by = "ID") %>%
  as.data.frame() 
}

ps_metadata <- calc_alpha_diversity(ps2)
```

### Distribution

#### Histogram
 
```{r, warning=F, message=F}
# To determine if diveristy is normally distributed
ggplot(ps_metadata, aes(x = Shannon)) + 
  geom_histogram() + 
  xlab("Alpha Diversity") + ylab("Count")
```

#### Test for normality.
 
```{r, warning=F, message=F}
shapiro.test(ps_metadata$Shannon)
```

### Plot 

#### Point

 - Use `plot_richness()` from *phyloseq*, which estimates alpha diversity metrics using *vegan* and plots them, taking standard *ggplot2* *geoms_* for the plot design.
 - use ps2 non-transformed data for alpha.
 
```{r, warning=F, message=F}
plot_richness(ps2, measures = c("Shannon", "Observed"), 
              color = "Probiotics", title = "") +
    geom_point(size = 3.5, alpha = 0.7) +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          panel.border = element_rect(colour = "grey", fill = NA, size = 1))+
  scale_color_manual(values = c("#440154FF", "#55C667FF"))+
  xlab("Infant Samples") +
  ylab("Alpha Diversity") 
```
#### Export plot
```{r, warning=F, message=F}
ggsave("outputs/shotgun/aplha_plot.png", dpi = 600 , height = 6, width = 7)
```

### Fit Model - Shannon

```{r,warning=F,message=F}
global <- glm(Shannon ~ Probiotics + Delivery + Diet, data = ps_metadata) 

global

car::Anova(global) %>% 
  as.data.frame(row.names = NULL)

emmeans(global, list(pairwise ~ Probiotics), adjust = "tukey") %>% 
  pairs() %>% 
  as.data.frame() 
```

```{r,warning=F,message=F,eval=F}
gof(global)
r.squaredGLMM(global)
```

### Fit Model - Observed

```{r,warning=F,message=F}
global <- glm(Observed ~ Probiotics + Delivery + Diet, data = ps_metadata) 

global

car::Anova(global) %>% 
  as.data.frame(row.names = NULL)

emmeans(global, list(pairwise ~ Probiotics), adjust = "tukey") %>% 
  pairs() %>% 
  as.data.frame() 
```

```{r,warning=F,message=F,eval=F}
gof(global)
r.squaredGLMM(global)
```
